Temperature Effects on Surface Sulfur Dioxide

### 2011-08-26 Background.tex
starting some more edits

A hydrated \suldiox~in an aqueous environment forms hydrogen bonds through the oxygens to nearby water hydrogens, or interacts via the sulfur atom with water oxygens. To further our analysis of the way in which \suldiox~coordinates its bonding to surface waters, those that lie in the topmost region of a gas/water interface, we adopt a naming scheme to denote the way in which the \suldiox~is hydrated by the surrounding waters. This naming scheme mimics a notational system developed during a previous study on water coordination by Buch et al,\cite{Buch 2005} and was subsequently used in other computational work.\cite{Walker2006b} In this naming system, a letter is used to designate the atom on a water molecule through which a hydrogen bond is formed to neighboring waters. Thus a bonding coordination of ``OOH'' designates two proton-acceptor bonding interactions through the water oxygen, and a single proton-donor bonding interaction through a hydrogen. More recently, Baer et al devised a nomenclature that explicitly enumerates the bonding to \suldiox~via the sulfur or oxygen atoms.\cite{Baer2010}

In this work we adopt Buch et al's nomenclature to \suldiox~in order to quantify hydrogen bonding through the acceptor \suldiox-oxygens, and the weaker bonding interactions from the \suldiox-sulfur to water-oxygens. Thus, an ``SOO'' coordinated \suldiox~molecule forms a single interaction through the sulfur atom to a neighboring water oxygen, and two hydrogen bonds through either a single \suldiox-oxygen, or distributed with one hydrogen bond on each of the \suldiox~oxygens. Analysis of the distribution of \suldiox~coordinations will give insight to how \suldiox~binds to the water surface.

In our analysis we defined bonds using the distance criteria of Baer et al. in determining the \suldiox~bonding coordinations.\cite{Baer2010} The bond-length definition is based on a set of distance criteria where a bonding interaction between a \wat-oxygen and \suldiox-sulfur is formed at a distance less than 3.5 \angs, and an \suldiox~oxygen hydrogen bond to a \wat-hydrogen is formed at a distance less than 2.2 \angs.

@@

\subsection {Graph Theoretical Details}
The optimized geometry of the \suldiox-hydrates suggests cyclic bonding structures, but it remains a different story entirely when a \suldiox~is placed in a dynamic environment such as in the course of molecular dynamics simulations of an aqueous surface. Geometry optimization shows the formation of these cyclic hydrate structures with two or three waters in a static vacuum. Do they also form in the course of a dynamic bonding process on a simulated water surface, where extended hydrating structures influence \suldiox~and water behavior? To study the formation and behavior of cyclic hydrate structures we employ graph theoretical techniques on MD trajectory data. Previous use of graphs in molecular computations were applied to finding stable arrangements of water clusters, ice, hydrogen bonding, extracting topological molecular properties, and cyclic structure studies.\cite{Anick2002, Huber2007, Radhakrishnan1991, Shi2005, Garcia2004, McDonald1998}

Here we briefly introduce graph theoretical concepts. They have been described well by others with varied application to cyclic structures.\cite{Tutte1984, Balakrishnan2000, Harary1973, Huber2007, Garcia2004, Dury2001} A graph consists of nodes, and edges that connect the nodes. A molecule can be represented with atoms as nodes, and edges for each intramolecular covalent bond connecting the atoms. The set of edges is then further expanded to include intermolecular interactions such as hydrogen bonds and other bonding bonding interactions. Edges may be assigned weights (i.e. bond lengths), types, and can be directional i.e. pointing towards a specific target from a source. A molecular system including all atoms, bonds, and interactions is thus fully described by a graph.

To detect cyclic structures in a graph a depth-first or breadth-first search (DFS and BFS, respectively) may be used.\cite{Knuth1997, Cormen 2001} A BFS is a recursive algorithm of queuing nodes and all neighboring nodes while performing a specified procedure on each visited node. This is easily performed on adjacency list or connectivity matrix data structures, iterating through nodes (i.e. atoms) of interest in the graph as starting points of the search. In BFS terminology, all nodes are colored during graph traversal to distinguish unvisited nodes (white), queued nodes (gray), and visited nodes (black). Using a BFS on a graph, cyclic structures are detected any time a ``gray target'' is encountered when queuing adjacent neighbors of a node. A benefit of BFS on a graph is the ability to determine the smallest cyclic structure in which a given node is a member. In the case of \suldiox~hydrate structures, beginning the BFS with the \suldiox~sulfur as the root node will discover cyclic bonding structures in order of size. Here we are only concerned with the smallest cyclic structure involving those waters in the first and second hydration shells around the \suldiox. Furthermore, it is possible to reconstruct a cycle's structure by finding its size (number of contributing atoms), and the number of unique waters in the cycle. This allows us to distinguish between various types of cyclic structures encountered.

Several arrangements of cyclic bonding structures are shown in Figure \ref{fig:cyclic-structures} for a \suldiox~molecule with three waters. Cycles with fewer or greater numbers of waters are also possible and encountered during MD. Cycle types \Rmnum{1}, \Rmnum{2}, \Rmnum{3} in Figure \ref{fig:cyclic-structures} are cyclic structures in which the \suldiox~is a member of the cycle. Types \Rmnum{4} and \Rmnum{5} do not involve the \suldiox~in the bonding cycle, but are commonly encountered as the smallest cycle formed near the \suldiox. Type \Rmnum{3} is of particular interest because the \suldiox~in the cycle has the most frequently occuring bonding coordination (``SO'', as shown later).

\begin{figure}[h!]
	\begin{center}
		\includegraphics[scale=1.0]{images/cycles/cycle-types-small.png}
		\caption{\suldiox~in involved in various cyclic structures encountered during MD simulations with water. The cartoons show the five types of cyclic structures, numbered for reference. The cyclic structures may involve any number of waters, but here each structure is shown with three waters. Type \Rmnum{3} is formed by a \suldiox~with the ``SO'' bonding coordination, which is the most dominant bonding coordination encountered.}
		\label{fig:cyclic-structures}
	\end{center}
\end{figure}

Baer et al. presented a detailed geometric and spectroscopic breakdown of type \Rmnum{3} cycles with two and three waters from their DFT calculations.\cite{Baer2010} Given the information of the number of waters, atoms, and bonds involved in the bonding cycles, we find that of the three-water type \Rmnum{3} cycles, there exist two structural varieties, shown in Figure \ref{fig:type-3-varieties}, that differ in the atomic contributions of the waters to the cyclic structure. Type \Rmnum{3}-A (shown in Figure \ref{fig:type-3-varieties}A) is arranged with each water contributing an OH bond to the structure of the cycle. Type \Rmnum{3}B involves a single water contributing an OH, whereas the other two waters contribute only the oxygen atom or the entire water molecule to the structure, respectively. This nuance of the type \Rmnum{3} structures involving three waters, and the overall distribution are presented in more detail later. We also show the distribution of cyclic structures encountered during MD simulations to further understand the behaviors of \suldiox-hydrates at the water surface.

\begin{figure}[h!]
	\begin{center}
		\includegraphics[scale=1.0]{images/cycles/triple-cycle-types-small.png}
		\caption{Type \Rmnum{3} cyclic structure (see Figure \ref{fig:cyclic-structures}) are found to occur in two varieties that are distinguished by the water atoms contributing to the cyclic structure. In ``type A'' the cycle is formed by an OH bond contributed by each water, and the SO bond of the \suldiox. ``type B'' involves an oxygen atom from one water, an OH bond from a second water, and all atoms of the third water.}
		\label{fig:type-3-varieties}
	\end{center}
\end{figure}

>>>>

A hydrated \suldiox~in an aqueous environment forms hydrogen bonds through the oxygens to nearby water hydrogens, or interacts via the sulfur atom with water-oxygens. To further our analysis of the way in which \suldiox~coordinates its bonding to surface waters, those that lie in the topmost region of a gas/water interface, we adopt a naming scheme to denote the way in which the \suldiox~is hydrated by the surrounding waters. This naming scheme mimics a notational system developed during a previous study on water coordination by Buch et al,\cite{Buch 2005} and was subsequently used in other computational work.\cite{Walker2006b} In this naming system, a letter is used to designate the atom on a water molecule through which a hydrogen bond is formed to neighboring waters. Thus a bonding coordination of ``OOH'' designates two proton-acceptor bonding interactions through the water-oxygen, and a single proton-donor bonding interaction through a hydrogen. More recently, Baer et al. devised a nomenclature that explicitly enumerates the bonding to \suldiox~via the sulfur or oxygen atoms.\cite{Baer2010}

In this work we adopt Buch et al's nomenclature to \suldiox~in order to quantify hydrogen bonding through the acceptor \suldiox-oxygens, and the weaker bonding interactions from the \suldiox-sulfur to water-oxygens. Thus, an ``SOO'' coordinated \suldiox~molecule forms a single interaction through the sulfur atom to a neighboring water-oxygen, and two hydrogen bonds through either a single \suldiox-oxygen, or distributed with one hydrogen bond on each of the \suldiox-oxygens. Analysis of the distribution of \suldiox~coordinations will give insight to how \suldiox~binds to the water surface.

In our analysis we defined bonds using the distance criteria of Baer et al. in determining the \suldiox~bonding coordinations.\cite{Baer2010} The bond-length definition is based on a set of distance criteria where a bonding interaction between a \wat-oxygen and \suldiox-sulfur is formed at a distance less than 3.5 \angs, and an \suldiox-oxygen hydrogen bond to a \wat-hydrogen is formed at a distance less than 2.2 \angs.

@@

\subsection {Graph Theoretical Details}
The optimized geometry of the \suldiox-hydrates suggests cyclic bonding structures, but it remains a different story entirely when a \suldiox~is placed in a dynamic environment such as in the course of molecular dynamics simulations of an aqueous surface. Geometry optimization shows the formation of these cyclic hydrate structures with two or three waters in a static vacuum. Do the cyclic structures also form in the course of a dynamic bonding process on a simulated water surface, where extended hydrating structures influence \suldiox~and water behavior? To study the formation and behavior of cyclic hydrate structures we employ graph theoretical techniques on MD trajectory data. Previous use of graphs in molecular computations were applied to finding stable arrangements of water clusters, ice, hydrogen bonding, extracting topological molecular properties, and cyclic structure studies.\cite{Anick2002, Huber2007, Radhakrishnan1991, Shi2005, Garcia2004, McDonald1998}

Here we briefly introduce graph theoretical concepts. They have been described well by others with varied application to cyclic structures.\cite{Tutte1984, Balakrishnan2000, Harary1973, Huber2007, Garcia2004, Dury2001} A graph consists of nodes, and edges that connect the nodes. A molecule can be represented with atoms as nodes, and edges for each intramolecular covalent bond connecting the atoms. The set of edges is then further expanded to include intermolecular interactions such as hydrogen bonds and other bonding bonding interactions. Edges may be assigned weights (i.e. bond lengths), types, and can be directional i.e. pointing towards a target node from a source node. A molecular system including all atoms, bonds, and interactions is thus fully described by a graph.

To detect cyclic structures in a graph a depth-first or breadth-first search (DFS and BFS, respectively) may be used.\cite{Knuth1997, Cormen 2001} A BFS is a recursive algorithm of queuing nodes and all neighboring nodes while performing a specified procedure on each visited node. This is easily performed on adjacency list or connectivity matrix data structures, iterating through nodes (i.e. atoms) of interest in the graph as starting points of the search. In BFS terminology, all nodes are colored during graph traversal to distinguish unvisited nodes (white), queued nodes (gray), and visited nodes (black). Using a BFS on a graph, cyclic structures are detected any time a ``gray target'' is encountered when queuing adjacent neighbors of a node. A benefit of BFS on a graph is the ability to determine the smallest cyclic structure containing a given node. In the case of \suldiox~hydrate structures, beginning the BFS with the \suldiox~sulfur as the starting, or root node or the search, will discover cyclic bonding structures in order of size. Here we are only concerned with the smallest cyclic structure involving those waters in the first and second hydration shells around the \suldiox. Furthermore, it is possible to reconstruct a cycle's structure by finding its size (number of contributing atoms), and the number of unique waters in the cycle. This allows us to distinguish between various types of cyclic structures encountered.

Several arrangements of cyclic bonding structures are shown in Figure \ref{fig:cyclic-structures} for a \suldiox~molecule with three waters. Cycles with fewer or greater numbers of waters are also possible and encountered during MD. Cycle types \Rmnum{1}, \Rmnum{2}, \Rmnum{3} in Figure \ref{fig:cyclic-structures} are cyclic structures in which the \suldiox~is a member of the cycle. Types \Rmnum{4} and \Rmnum{5} do not involve the \suldiox~in the bonding cycle, but are commonly encountered as the smallest cycle formed near the \suldiox. Type \Rmnum{3} is of particular interest because the \suldiox~in this cycle has the most frequently occuring bonding coordination (``SO'', as shown later).

\begin{figure}[h!]
	\begin{center}
		\includegraphics[scale=1.0]{images/cycles/cycle-types-small.png}
		\caption{\suldiox~in various cyclic structures encountered during MD simulations with water. The cartoons show the five types of cyclic structures, numbered for reference. The cyclic structures may involve any number of waters, but here each structure is shown with three waters. Type \Rmnum{3} is formed by a \suldiox~with the ``SO'' bonding coordination, which is the most dominant bonding coordination encountered.}
		\label{fig:cyclic-structures}
	\end{center}
\end{figure}

Baer et al. presented a detailed geometric and spectroscopic breakdown of type \Rmnum{3} cycles with two and three waters from their DFT calculations.\cite{Baer2010} Given the information of the number of waters, atoms, and bonds involved in the bonding cycles, we find that of the three-water type \Rmnum{3} cycles, there exist two structural varieties, shown in Figure \ref{fig:type-3-varieties}, that differ in the set of water atoms involved in the cyclic structure. Type \Rmnum{3}-A (shown in Figure \ref{fig:type-3-varieties}A) is arranged with each water contributing an OH bond to the structure of the cycle. Type \Rmnum{3}B involves a single water contributing an OH, whereas the other two waters contribute only the oxygen atom or the entire water molecule to the structure, respectively. This nuance of the type \Rmnum{3} structures involving three waters, and the overall distribution are presented in more detail later. We also show the distribution of cyclic structures encountered during MD simulations to further understand the behaviors of \suldiox-hydrates at the water surface.

\begin{figure}[h!]
	\begin{center}
		\includegraphics[scale=1.0]{images/cycles/triple-cycle-types-small.png}
		\caption{Type \Rmnum{3} cyclic structures (see Figure \ref{fig:cyclic-structures}) are found to occur in two varieties that are distinguished by the water atoms contributing to the cyclic structure. In ``type A'' the cycle is formed by an OH bond contributed by each water, and the SO bond of the \suldiox. ``type B'' involves an oxygen atom from one water, an OH bond from a second water, and all atoms of the third water.}
		\label{fig:type-3-varieties}
	\end{center}
\end{figure}

### 2011-08-26 ComputationalMethods.tex
starting some more edits

On-the-fly ab initio molecular dynamics simulations were performed with the QUICKSTEP package, which is an implementation of the Gaussian plane wave method using the Kohn-Sham formulation of density functional theory (DFT).~\cite{VandeVondele2005} The Kohn-Sham orbitals are expanded using a linear combination of atom-centered Gaussian-type orbital functions. The electronic charge density was desccribed using an auxilliary basis set of plane waves. Energies and forces from on-the-fly simulation sampling of the Born-Oppenheimer surface were calculated for each MD step using the Gaussian DZVP basis set, the exchange-correlation functional of Becke, Lee, Yang, and Parr (BLYP)~\cite{LEE1988}, and the atomic pseudo potentials of the Goedecker, Teter, and Hutter type.~\cite{Goedecker1996} A simulation timestep of 1.0 fs was used, with a Nose-Hoover thermostat set at 273K and 300K for the cold and hot simulations, respectively. These computational parameters were verified to yield a reasonable description of bulk room temperature water when simulating a neat-water system.

Initially, 10 equilibrated boxes of side-lengths 10.0\angs~, with 36 randomly packed water molecules were used. 5 of the boxes were used for each of the cold and hot simulations. A sulfur dioxide molecule was randomly placed onto the surface within 2.5\angs~of a water molecule centrally located above the waters in the z-axis. A copy of the initial system cubes were then expanded along one axis (z-axis) to 25\angs. The system energy was minimized. Subsequently, the system was equilibrated for 1 ns in canonical ensemble (NVT) conditions. Periodic boundaries were set on the two short axes to form an infinite slab. The equilibrated systems were then simulated for a further 20 ps in the microcanonical ensemble (NVE), with trajectory snapshots recorded every 1.0 fs. The initial 1 ns equilibration trajectory was not included in the final analysis. This resulted in 20,000 time steps of system trajectory for analysis in the sets of hot and cold replicas of the system.

>>>>

On-the-fly ab initio molecular dynamics simulations were performed with the QUICKSTEP package, which is an implementation of the Gaussian plane wave method using the Kohn-Sham formulation of density functional theory (DFT).~\cite{VandeVondele2005} The Kohn-Sham orbitals are expanded using a linear combination of atom-centered Gaussian-type orbital functions. The electronic charge density was described using an auxiliary basis set of plane waves. Energies and forces from on-the-fly simulation sampling of the Born-Oppenheimer surface were calculated for each MD step using the Gaussian DZVP basis set, the exchange-correlation functional of Becke, Lee, Yang, and Parr (BLYP)~\cite{LEE1988}, and the atomic pseudo-potentials of the Goedecker, Teter, and Hutter type.~\cite{Goedecker1996} A simulation timestep of 1 fs was used, with a Nose-Hoover thermostat set at 273K and 300K for the cold and hot simulations, respectively. These computational parameters were verified to yield a reasonable description of bulk room temperature water when simulating a neat-water system.

Initially, 10 equilibrated boxes of side-lengths 10.0\angs~, with 36 randomly packed water molecules were used. Five of the boxes were used for each of the cold and hot simulations. A sulfur dioxide molecule was randomly placed onto the surface within 2.5\angs~of a water molecule centrally located above the waters in the z-axis. A copy of the initial system cubes were then expanded along one axis (z-axis) to 25\angs. The system energy was minimized. Subsequently, the system was equilibrated for 1 ns in canonical ensemble (NVT) conditions. Periodic boundaries were set on the two short axes to form an infinite slab. The equilibrated systems were then simulated for a further 20 ps in the microcanonical ensemble (NVE), with trajectory snapshots recorded every 1 fs. The initial 1 ns equilibration trajectory was not included in the final analysis. This resulted in 20,000 time steps of system trajectory for analysis in the sets of hot and cold replicas of the system.

### 2011-08-26 Coordination-Results.tex
starting some more edits

The bonding coordination of the surface \suldiox~was determined at each timestep of the simulations. Figure \ref{fig:bonding-coordinations} shows the distribution of bonding coordinations of the surface \suldiox, by percentage of all bonding coordinations encountered for both the cold (blue) and hot (red) trajectories. A first visual inspection reveals several trends. Clearly the ``SO'' coordination is the most populous at both temperatures. The second and third most populated coordinations are the ``S'' and ``SOO'', however the distributions differ between temperatures. In cold simulations the ``S'' and ``SOO'' coordinations occur nearly equally. The hotter temperature simulation shifts the distribution such that the ``S'' occurs 5\% less frequently than in the cold, and the ``SOO'' occurs nearly 10\% more often. The distribution of bonding coordinations in the hot temperature has a clear first and second most frequent coordination: ``SO'' and ``SOO'', respectively. These results for the hot (room temperature) system coincide with those of the previous single-temperature simulation study by Baer et al. at a similar temperature.\cite{Baer2010}

@@

	Several conclusions about the bonding behavior of \suldiox~to surface waters stem from this distribution of coordinations. Clearly the cold \suldiox~spends more time than the hot \suldiox, 13\% and 3\%, respectively, completely unbound from the surface waters. This does not necessarily imply a complete desorption into the gas phase, but only a brief sojourn away from the waters, with all interactions and bond lengths longer than the cutoff criteria used for the analysis. Furthermore, the most frequently occuring bonding coordinations are ``S'',``SO'', and ``SOO'', with ``SO'' being the most populated at both temperatures. Baer et al. also concluded that these three coordinations were the most frequent for a room temperature simulation, and specifically identified the ``SO'' and ``SOO'' as most common in their study.\cite{Baer2010}

	Looking closer at the coordination types it is notable that coordinations lacking any sulfur interactions (i.e. ``O'', ``OO'', etc.) represent the least frequently formed bonding coordinations. A bonding coordination with at least a single sulfur interaction is clearly favored over \suldiox~``oxygen-only'' bonding to waters. In our previous classical simulations of \suldiox~on water we concluded that during adsorption and throughout the interface, the \suldiox~orients so that its sulfur tends to the \wat~bulk side of the interface.\cite{Shamay2011} The coordination distributions here support the idea that binding through the sulfur is preferable, and to the extent that a non-sulfur coordination is rarely formed during the course of all our simulations.

>>>>

The bonding coordination of the surface \suldiox~was determined at each timestep of the simulations. Figure \ref{fig:bonding-coordinations} shows the distribution of bonding coordinations of the surface \suldiox, by percentage of all bonding coordinations encountered for both the cold (blue) and hot (red) trajectories. A first visual inspection reveals several trends. Clearly the ``SO'' coordination is the most populous at both temperatures. The second and third most populated coordinations are the ``S'' and ``SOO'', however their distributions differ between temperatures. In cold simulations the ``S'' and ``SOO'' coordinations occur nearly equally. The hotter temperature simulation shifts the distribution such that the ``S'' occurs 5\% less frequently than in the cold, and the ``SOO'' occurs nearly 10\% more often. The distribution of bonding coordinations in the hot temperature has a clear first and second most frequent coordination: ``SO'' and ``SOO'', respectively. These results for the hot (room temperature) system coincide with those of the previous single-temperature simulation study by Baer et al. at a similar temperature.\cite{Baer2010}

@@

	Several conclusions about the bonding behavior of \suldiox~to surface waters stem from this distribution of coordinations. Clearly the cold \suldiox~spends more time than the hot \suldiox, 13\% versus 3\%, respectively, completely unbound from the surface waters. This does not necessarily imply a complete desorption into the gas phase, but only a brief sojourn away from the waters, with all interactions and bond lengths longer than the cutoff criteria used for the analysis. Furthermore, the most frequently occuring bonding coordinations are ``S'',``SO'', and ``SOO'', with ``SO'' being the most populated at both temperatures. Baer et al. also concluded that these three coordinations were the most frequent for a room temperature simulation, and specifically identified the ``SO'' and ``SOO'' as most common in their study.\cite{Baer2010}

	Looking closer at the coordination types it is notable that coordinations lacking any sulfur interactions (i.e. ``O'', ``OO'', etc.) represent the least frequently formed. A bonding coordination with at least a single sulfur interaction is clearly favored over \suldiox~``oxygen-only'' bonding to waters. In our previous classical simulations of \suldiox~on water we concluded that during adsorption and throughout the interface, the \suldiox~orients so that its sulfur tends to the \wat~bulk side of the interface.\cite{Shamay2011} The coordination distributions here support the idea that binding through the sulfur is preferable, and to the extent that a non-sulfur coordination is rarely formed during the course of all our simulations.
